---
title: "Lab2_special task"
author: "Andreas"
date: "10 Dec 2018"
output: 
  pdf_document:
    toc: True
---


##Special Task 1

```{r,echo=FALSE}

LDA <- function(x, y) {
  #indexes of 2 classes
  group1_index <- which( y == "Male" )
  group2_index <- which( y == "Female" )
  #priors probs of classes
  p1 <- length(group1_index) / length(y)
  p2 <- length(group2_index) / length(y)
  prior_probs<-cbind(p1,p2)
  #means of the classes
  mean_class1 <- colMeans(x[group1_index, ])
  mean_class2 <- colMeans(x[group2_index, ])
  group_centers<-cbind(mean_class1,mean_class2)
  #subtract the mean of the group from each group
  x1 = as.matrix(sweep(x[group1_index, ], 2, mean_class1, "-"))
  x2 = as.matrix(sweep(x[group2_index, ], 2, mean_class2, "-"))
  #covariance matices 
  C1 = cov(x1)     
  C2 = cov(x2)      
  #within-class matrix
  S_w = (C1 + C2 )
  #means of global means
  mean_global<-colMeans(x)
  #between-class matrix 
  S1<-(mean_class1-mean_global)%*%t(mean_class1-mean_global)
  S2<-(mean_class2-mean_global)%*%t(mean_class2-mean_global)
  S_b<-length(group1_index)*S1+length(group2_index)*S2
  #eigen vectors ,slope,intercept of decision equation
  ev <- eigen(solve(S_w) %*% S_b)$vectors
  means<-rbind(mean_class2,mean_class1) #rbind means of 2 classes
  rownames(means)<-c("Female","Male")
  gmean<-prior_probs%*%means
  const <- drop(gmean%*%ev)
  lda_slope <- - ev[1,1] / ev[2,1]    #slope of decision line
  lda_intercept <- (const/ev[2,1])[1]  #intercept of decision line
  

  #discriminat function of male
  a0_m<- -0.5*(mean_class1%*%solve(S_w)%*%mean_class1)
  a1_m<- solve(S_w)%*%mean_class1 
  a0_f<--0.5*(mean_class2%*%solve(S_w)%*%mean_class2)
  a1_f<- solve(S_w)%*%mean_class2 

  
  a0<-log(p1/p2)-0.5*((mean_class1+mean_class2)%*%solve(S_w)%*%(mean_class2-mean_class1))
  a1<-t((mean_class2-mean_class1))%*%solve(S_w) #%*%t(as.matrix(x))
  
  
  
  #make predictions if w*x+C>0-->Male else Female
  w<-as.matrix(t(solve(S_w)%*%(mean_class1-mean_class2)))%*%t(as.matrix(x))
  c<-log(p1/p2)-0.5*(mean_class1%*%solve(S_w)%*%mean_class1)+0.5*(mean_class2%*%solve(S_w)%*%mean_class2)
  preds2<-ifelse(apply(w,2,function(x){x+c})>0,"Male","Female")
  
  
  
  return(list("predictions"=preds2,"slope"=lda_slope,"intercept"=lda_intercept,"Group_means"=means,
              "discriminant_function_Male"=c("inter"=a0_m,"coeffs"=a1_m),
              "discriminant_function_Female"=c("inter"=a0_f,"coeffs"=a1_f),
              "equation of decision bountary"=c("inter"=a0,"coeffs"=a1),"C1"=C1))
  
}


```



###1
```{r,echo=FALSE}
crab<-read.csv("australian-crabs.csv",sep=",",header = T)

x<-crab[,5:6]
y<-crab[,2]


L<-LDA(x,y)


cat("The equation of linear dicriminat function for Male is:\n",
    L$discriminant_function_Male)
cat("\n")
cat("The equation of linear dicriminat function for Female is:\n",
    L$discriminant_function_Female)
cat("\n")
cat("The equation for decision boundary is :\n",
    "slope=",L$slope,"intercept"=L$intercept)
```
As we can see the equation of linear discriminat function for class Male is : $-5.935135+1.282926*RW- 0.1069072*CL$ and for the Female class is :
$-10.86781+4.124349*RW-1.080659*CL$.
Finally,the equation for decision boundary is $-4.932676+2.841423*RW-1.080659*CL$ and the line for decision line is :  
$y=2.928015-5.065639*x$




###2

```{r,echo=F}
paste0("The accuracy of the model is :",mean(y==L$predictions))

plot(x[,1],x[,2],col=as.factor(L$predictions),xlab = "RW",ylab = "CL",
     main = "CL~RW colored by sex and decision boundary",pch=16)
abline(L$intercept,L$slope,col="green",lty=2, lwd=3)

```

In the above plot we can see the decision boundary for our model with the green dashed line.Also calculating the accuracy of the model we can see that it is 
96.5% which is very high and our model does a very good job and our predictions are very accurate.The model depends on the prior probabilities and in our case 
we have equaly sized classes and the linear line is able to seperate the classes very accurate.


##Special Task2



